---
title: "Moldova PRO-EU Referendum"
author:
  - name: Sergiu Lipcean
    url: https://sergiulipcean.github.io/
    orcid: 0000-0002-8944-6505
date: "2024-11-07"
categories: [referendum, european union, constitutional amendment, democracy, plotly] # self-defined categories
draft: false # setting this to `true` will prevent your post from appearing on your listing page until you're ready!
link-citations: true
zotero: true
linkcolor: blue
urlcolor: blue
---

```{=html}
<style type = text/css>
  body{line-height: 1.5em; 
  font-size: 14pt;
  font-family: Trebuchet MS;
  text-align: justify;}
</style>
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.width = 10, fig.height = 10)
```

```{r}
# Download R packages 
library(tidyverse)
library(plotly)
library(webshot)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
library(scales)
```

```{r}

#| label: data-management

referendum_data <- readRDS("referendum_map_data.rds") %>%
  mutate(diff_shares = diff_shares*100)
```

```{r}

#| fig-height: 8
#| fig-width: 6

# Create the subtitle as a separate annotation
subtitle_annotation <- list(
  text = "District-Level Results",
  xref = "paper",
  yref = "paper",
  xanchor = "center",
  yanchor = "top",
  x= 0.55,
  y = 1,
  sizex = 1,
  sizey = 1,
  showarrow = FALSE,
  font = list(size = 16)
)

# Create the caption annotation
caption_annotation <- list(
  text = "Data Source: Own elaboration based on Central Electoral Commission official results",
  xref = "paper",
  yref = "paper",
  xanchor = "right",
  yanchor = "bottom",
  x = 1,
  y = 0,
  sizex = 1,
  sizey = 1,
  showarrow = FALSE,
  font = list(size = 12)
)

# Create the plotly choropleth map
referendum_map <- plot_ly() %>%
  add_sf(
    data = referendum_data,
    color = ~diff_shares,
    colors = "plasma",
    split = ~rayon,
    showlegend = FALSE,
    text = ~paste("District:", rayon, 
                  "<br>Difference:", diff_shares,"%"),
    hoverinfo = "text",
    hoveron = "fills",
    opacity = 1,
    line = list(
      width = 0.5)) %>%
  colorbar(title = "% Difference") %>%
  layout(
    title = "",
    annotations = list(subtitle_annotation, caption_annotation),
    geo = list(
      showframe = FALSE,
      projection = list(type = "mercator"),
      fitbounds = "locations"
    ),
    colorbar = list(x = 0.7, # Adjust the x position to move the colorbar horizontally
                y = 0.5, # Adjust the y position to move the colorbar vertically
                len = 0.7 # Adjust the length of the colorbar
    )
   
  )
referendum_map

```

```{=html}
<style type = text/css>
  body{line-height: 1em; 
  font-size: 12pt;
  font-family: Trebuchet MS;}
</style>
```

<br> Note: The plot displays the percentage difference at the district (rayon) level between votes cast PRO and AGAINST amending the constitution for joining the European Union (Referendum was held on 20 October 2024). Use the mouse to hover over the plot for results in each district. Referendum question: Do you support the amendment of the Constitution in order for the Republic of Moldova to join the European Union?

