---
title: "Access to subsidies, financial transparency & abuse of state resources in elections by democracy and income levels"
author:
  - name: Sergiu Lipcean
    url: https://sergiulipcean.github.io/
    orcid: 0000-0002-8944-6505
date: "2022-09-14"
categories: [public funding access, transparency, elections, abuse of state resources, democracy, income, plotly] # self-defined categories
draft: false # setting this to `true` will prevent your post from appearing on your listing page until you're ready!
link-citations: true
csl: apa.csl
zotero: true
linkcolor: blue
urlcolor: blue
bibliography: references.bib
---

```{=html}
<style type = text/css>
  body{line-height: 2em; 
  font-size: 14pt;
  font-family: Trebuchet MS;}
</style>
```
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.width = 10, fig.height = 10) 
```

```{r}
# Download R packages 
library(tidyverse)
library(plotly)
library(haven)
#library(car)
library(rgl)
library(webshot)
```

```{r}
pei <- read_dta("PEI election-level data (PEI_8.0).dta")
pei <- pei %>%
  mutate(gdp_pc_ppp = round(gdp_pc_ppp,1),
         subsidies = round(subsidies, 2),
         donations = round(donations, 2),
         rich = round(rich,2),
         resources = round(resources,2), 
         fh_score = fhcategory + 1,
         fh_score = factor(fh_score, labels = c("Not Free", "Partly Free", "Free")))
```

```{r}
m <- list(l = 0,  r = 0, b = 0, t = 0, pad = 4)

axx <- list(title = "Access to subsidies")
axy <- list(title = "Abuse of state resources")
axz <- list(title = "Financial transparency")
my_legend <- list(title=list(text='Democracy'))

pei %>%
  plot_ly(x = ~subsidies, y = ~resources, z = ~accounts, size = ~gdp_pc_ppp,
          marker = list(symbol = 'circle', opacity = 0.5, sizemode = 'diameter'), sizes = c(4, 40),
          color = ~fh_score,
          colors = c("red4", "yellow2", "darkgreen"),
          hoverinfo = 'text',
          text = ~paste("Country: ", ISO, "<br>",
                        "Subsidies: ", subsidies, "<br>",
                        "Donations: ", donations, "<br>",
                        "Abuse : ", resources, "<br>", 
                        "GDP per capita", gdp_pc_ppp)) %>%
      # Axes labels and title:
  layout(scene = list(xaxis = axx, yaxis = axy, zaxis = axz),
         legend = list(title=list(text='Democracy'), 
                       font = list(size = 14),
                       y = 0.5),
         margin = m)

```

```{=html}
<style type = text/css>
  body{line-height: 1em; 
  font-size: 12pt;
  font-family: Trebuchet MS;}
</style>
```
<br> Note: The plot depicts the relationship between the access (higher values - fairer access) to state subsidies for campaigning, publication of financial accounts (higher values - more transparency), on the one hand, and the abuse of state resources (higher values - more abuse), on the other hand. Additionally, it breaks down this relationship by the level of democratic development (colour) and GDP per capita (larger points - higher GDP). Overall, democracy and wealth are associated with fairer access to subsidies and more transparency of financial accounts than partly free and non-free regimes. Data reflects expert opinions. Use the mouse to rotate/spin, zoom in, and hover over the plot for more details. Data source: [@garnett2022]

::: {#references}
:::
