---
title: "Relationship between fair access to public funding of parties and campaign donations and the capacity of the rich to buy elections by the level of electoral integrity"
author:
  - name: Sergiu Lipcean
    url: https://sergiulipcean.github.io/
    orcid: 0000-0002-8944-6505
date: "2021-11-23"
categories: [public funding of parties, donations, elections, electoral integrity, plotly] # self-defined categories
draft: false # setting this to `true` will prevent your post from appearing on your listing page until you're ready!
link-citations: true
csl: apa.csl
zotero: true
linkcolor: blue
urlcolor: blue
bibliography: references.bib
---

```{=html}
<style type = text/css>
  body{line-height: 2em; 
  font-size: 14pt;
  font-family: Trebuchet MS;}
</style>
```
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.width = 10, fig.height = 10) 
```

```{r}
# Download R packages 
library(tidyverse)
library(haven)
#library(scales) 
library(car)
library(rgl)
library(plotly)
library(webshot)
```

```{r}
# Preprocesiing 

pei <- read_dta("PEI election-level data (PEI_8.0).dta")

pei <- pei %>%
  mutate(gdp_pc_ppp_ln = log(gdp_pc_ppp),
         PEItype_fct = factor(PEItype, labels = c("Very Low", "Low", "Medium", "High", "Very High")),
         fh_score = fhcategory + 1,
         fh_score = factor(fh_score, labels = c("Not Free", "Partly Free", "Free"))) %>%
  sample_n(479)

x_values <- pei$subsidies %>% round(3)
y_values <- pei$donations %>% round(3)
z_values <- pei$rich %>% round(3)

# Define regression plane -------------------------------------------------
# Construct x and y grid elements
x_grid <- seq(from = min(x_values, na.rm = TRUE), to = max(x_values, na.rm = TRUE), length = 50)
y_grid <- seq(from = min(y_values, na.rm = TRUE), to = max(y_values, na.rm = TRUE), length = 50)

# Construct z grid by computing
# 1) fitted beta coefficients
# 2) fitted values of outer product of x_grid and y_grid
# 3) extracting z_grid (matrix needs to be of specific dimensions)
beta_hat <- pei %>% 
  #drop_na(rich, subsidies, donations) %>%
  lm(rich ~ subsidies + donations, data = .) %>% coef()

fitted_values <- crossing(y_grid, x_grid) %>% 
  mutate(z_grid = beta_hat[1] + beta_hat[3]*y_grid + beta_hat[2]*x_grid)

z_grid <- fitted_values %>% 
  pull(z_grid) %>%
  matrix(nrow = length(x_grid)) %>%
  t()

# Define text element for each point in plane
text_grid <- fitted_values %>% 
  pull(z_grid) %>%
  round(3) %>% 
  as.character() %>% 
  paste("rich: ", ., sep = "") %>% 
  matrix(nrow = length(x_grid)) %>%
  t()

```

```{r}

m <- list(l = 0,  r = 0, b = 0, t = 0, pad = 4)

# Plot using plotly
pei %>%
  plot_ly() %>%
  add_markers(x = x_values, y = y_values, z = z_values, marker = list(size = 5, opacity = 0.7), 
          
          symbol = ~PEItype_fct,
          symbols = c("circle", "square", "x", "diamond", "cross"),
          color = ~PEItype_fct,
          colors = "viridis",
          #colors = hcl.colors(20, "Plasma"),
          reversescale = TRUE,
          hoverinfo = 'text', 
          text = ~paste("Country: ", ISO, "<br>",
      "Rich: ", z_values, "<br>",
      "Donations: ", y_values, "<br>",
      "Subsidies: ", x_values 
    )) %>%
  
  # Regression plane:
    add_surface(
    x = x_grid,
    y = y_grid,
    z = z_grid,
    opacity = 0.8,
    hoverinfo = 'text',
    text = text_grid
  ) %>%
  # Axes labels and title:
  layout(
    scene = list(
      zaxis = list(title = "y: Rich"),
      yaxis = list(title = "x2: Donations"),
      xaxis = list(title = "x1: Subsidies")),
    legend = list(title=list(text='Electoral<br>integrity'), font = list(size = 14), y = 0.4),
    margin = m
  ) %>%
    colorbar(title=list(text='Likelyhood<br>of the rich<br>to buy elections'), tickfont=list(size=14), y = 0.9)
#legend = list(font = list(size = 30))

```

```{=html}
<style type = text/css>
  body{line-height: 1em; 
  font-size: 10pt;"}
</style>
```

<br>
Note: The plot illustrates the relationship between fair access to state funding of parties and campaign donations (higher values - fairer access), and the capacity of the rich to buy elections (higher values - rich more influential) by the overall level of electoral integrity. It shows that a fairer access to state funding and private contributions is associated with a higher level of electoral integrity and lower ability of the rich to buy elections. Data reflects expert opinions. Use the mouse to rotate/spin, zoom in, and hover over the plot for more details. Data source: [@garnett2022]

::: {#references}
:::
