---
title: "Party registration requirements across post-communist regimes 1991-2020"
author:
  - name: Sergiu Lipcean
    url: https://sergiulipcean.github.io/
    orcid: 0000-0002-8944-6505
date: "2024-07-20"
categories: [party registration, freedom of association, democracy, plotly] # self-defined categories
draft: false # setting this to `true` will prevent your post from appearing on your listing page until you're ready!
link-citations: true
csl: apa.csl
zotero: true
linkcolor: blue
urlcolor: blue
bibliography: references.bib
---

```{=html}
<style type = text/css>
  body{line-height: 1.5em; 
  font-size: 14pt;
  font-family: Trebuchet MS;
  text-align: justify;}
</style>
```
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.width = 10, fig.height = 10)
```

```{r}
# Download R packages 
library(tidyverse)
library(openxlsx)
library(plotly)
library(rgl)
library(scales)
library(webshot)
```

```{r}

#| label: data-management

teritorial_distrib <- read.xlsx("openess2.xlsx", sheet = "ter") %>%
  mutate(country_year = str_c(iso, year, sep = "-"),
         country = factor(country)) %>%
  ungroup() %>%
  dplyr::select(iso, year, country_year, territorial, branches)

openess <- read.xlsx("openess2.xlsx") %>%
  mutate(country = factor(country)) %>%
  group_by(country) %>%
  mutate(restrictive_prc = round(number/regist_v*100, 4)) %>%
  left_join(teritorial_distrib, by = c("iso", "year")) %>%
  mutate(across(.cols= territorial:branches, ~ifelse(is.na(.x), 0, .x))) %>%
  drop_na(restrictive_prc)
```

```{r}

#| fig-height: 5
#| fig-width: 6

### Custom colours
my_colurs <- c("#E41A1C", "#377EB8", "#4DAF4A", "#984EA3", "#FF7F00", "#FFFF33", 
            "#A65628", "#F781BF", "#999999", "#66C2A5", "#FC8D62", "#8DA0CB", 
            "#E78AC3", "#A6D854", "#FFD92F", "#E5C494", "#B3B3B3", "#8DD3C7",
            "#FFFFB3", "#BEBADA", "#FB8072", "#80B1D3", "#FDB462", "#B3DE69",
            "#FCCDE5", "#D9D9D9", "#BC80BD")

### Custom plot margins
m <- list(l = 0,  r = 0, b = 0, t = 0, pad = 4)

### Draw scatterplot
plotly_plot <- plot_ly(
  openess,
  x = ~regist_v / 1e6,  # Convert registered voters to millions
  y = ~number,
  z = ~territorial + 1,  
  color = ~country,
  size = ~restrictive_prc,
  hoverinfo = 'text',
  text = ~paste("Country-year: ", cntr_year, "<br>",
                "% voters: ", restrictive_prc),
  type = 'scatter3d',  # Use 3D scatter plot type
  mode = 'markers',
  marker = list(
    symbol = 'circle',
    opacity = 0.6,
    sizemode = 'diameter'
  ),
  sizes = c(7, 70),  # Set size range directly
  colors = my_colurs
) %>%
  # Customize the layout for axes, legend, and annotations
  layout(
    title = "",
    scene = list(
      xaxis = list(
        title = "Registered voters in millions (log scale)",
        type = "log",
        tickvals = c(0, 1, 3, 8, 15, 30, 60, 100)
      ),
      yaxis = list(
        title = "Number of required signatures (log scale)",
        type = "log",
        tickvals = c(0, 5, 10, 30, 100, 300, 1000, 3000, 10000, 50000)
      ),
      zaxis = list(
        title = "Territorial distribution (No signatures)", 
        type = "log",
        tickvals = c(1, 10, 20, 50, 100, 300, 700)
      )
    ),
    margin = m,
    showlegend = TRUE,
    legend = list(title = list(text = "% Voters"))
    )

# Display the plot
plotly_plot

```

```{=html}
<style type = text/css>
  body{line-height: 1em; 
  font-size: 12pt;
  font-family: Trebuchet MS;}
</style>
```

<br> Note: The plot depicts the relationship between the size of electoral market measured as the number of registered voters, the number of voters' signatures required to register a new political party and the territorial distribution of new party members across administrative-territorial units within each country. The size of the bubbles, represents the percentage of voters from which to recruit a single member and reflects,  the effort required by newly aspiring political organisations to recruit potential members from the voters' pool. The larger the bubbles, the more burdensome are party registration provisions, thus demanding more efforts on behalf of new political parties. Additionally, the higher the number of signatures for the "Territorial distribution" axis, the more difficult for new parties to register. Use the mouse to rotate/spin, zoom in and out, and hover over the plot for more details. Data source: Own data collected from party laws and regulations of the respective countries over three decades.

::: {#references}
:::
