---
title: "Direct public funding of political parties: between proxy measures and hard data (2)"
description: "In this blogpost I analyse the development of public funding regimes in postcommunist polities"
author:
  - name: Sergiu Lipcean
    url: https://sergiulipcean.github.io/
    orcid: 0000-0002-8944-6505
date: "2021-09-16"
categories: [public funding of parties, hard data, post-communist regimes] # self-defined categories
draft: false # setting this to `true` will prevent your post from appearing on your listing page until you're ready!
link-citations: true
csl: apa.csl
zotero: true
linkcolor: blue
urlcolor: blue
bibliography: references.bib
resources:
  - "dpf_dataset.csv"
  - "dpf_dependence.csv"
---

```{=html}
<style type = text/css>
  body {line-height: 1.5em;
  font-size: 14pt;
  font-family: Trebuchet MS;
  text-align: justify;
  }
</style>
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r}
# Download R packages 
library(tidyverse)
library(scales) 
library(ggthemes) 
library(ggrepel) 
library(ggpubr) 
library(svglite) 
library(cowplot) 
library(ggsci)
```

```{r}
####################################
### Import the dataset from the working directory 
dpf_dataset <- read_csv("dpf_dataset.csv", show_col_types = FALSE)
```

```{r}
custom_theme <- theme_few(base_size = 12, base_family = "calibri") +
  theme(panel.border = element_rect(color = "gray60"),
        panel.grid.major = element_line(color = "grey90", size= 0.1))
```

|       In the previous blogpost, I argued that binary indicators of party funding such presence vs absence of public funding, donation or spending caps are poor proxies to use in research assessing the impact of financing regulations on political competition, corruption, or other phenomena of interest. The same applies to aggregate indexes of party and campaign funding, which combine different regulatory dimensions (donation caps + spending caps + public funding + transparency) and derive a composite measure through their addition or some alternative procedure. Since comparative cross-national data of party and campaign funding is difficult to obtain, there is a tendency to use expert-based evaluations of different aspects, such as the significance of public funding for the election campaign of parties and candidates or the fairness in the distribution of campaign subsidies. Yet, the critical challenge these measures face is their validity -- how accurately expert assessments capture the actual situation. The analysis of two such indexes against real data reveals a rather low validity. @fig-experts presents the results.[^1] In the left-hand panel, I correlate my data on the public funding level [@lipcean2021] with the expert scores average for the question from the Varieties of Democracy dataset "Is significant public inancing available for parties' and/or candidates' campaigns for national office?" (ordinal scale ranging from 0 (no subsidies) to 4 (significant subsidies)) [@coppedge2020; @coppedge2021]. In the right-hand panel, I correlate my data on the legal requirements to get access to campaign subsidies with the expert scores average from the "Electoral Integrity Project" for the question "Parties/candidates had equitable access to public political subsidies?" (ordinal scale ranging from 1 (strong disagreement) to 5 (strong agreement)) [@norris2019; @norris2019a]. As one may notice, the relationship between actual indicators of public funding and expert assessments produce a considerable mismatch, indicating that expert judgements must be treated cautiously in research on political financing.

[^1]: You can replicate @fig-experts by using this data [{{< fa file-csv >}}](dpf_dataset.csv)`dpf_dataset.csv` along with the R code.

```{r}
#| label: fig-experts
#| fig-cap: "Relationship between expert scores, subsidy level and payout threshold"
#| fig-width: 8
#| fig-height: 4
expert_amount <- dpf_dataset %>% 
  ggplot(aes(x = dpfv_stat_usd, y = v2elpubfin, 
             color = country_id)) +
  geom_jitter(width = 0.3, height = 0.2, shape = 1, size = 2, alpha = 0.6) +
  scale_color_igv(name = "", palette = "default") +
  guides(color = guide_legend(ncol=1)) +
  custom_theme +
  theme(plot.title = element_text(size = 12),
        legend.text = element_text(size = 8),
        legend.box.just = "top",
        legend.key.size = unit(0.6, 'lines')) +
  labs(title = "A: Expert scores vs. subsidy level", y = "Low-High subsidy significance: expert scores", x = "Actual subsidy per vote USD")

expert_access <- dpf_dataset %>%
  mutate(labels_access_el = str_c(country_id, year, sep = "-")) %>% ### Create country-year labels
  filter(access_elections < 10) %>%
  ggplot(aes(x = access_elections, y = subsidies_eip)) +
  geom_point(aes(color = country_id),shape = 1, size = 2) +
  scale_color_igv(palette = "default") +
  geom_text_repel(aes(label = labels_access_el, color = country_id), size = 3, family = "serif") +
  scale_x_continuous(breaks = seq(0,6,1)) +
  custom_theme +
  theme(plot.title = element_text(size = 12),
        legend.position = "none") +
  labs(title = "B: Expert scores vs. access threshold", y = "Disagree vs. Agree: Equitable access to subsidies", x = "Access to subsidies (% of votes/seats)")

plot_grid(expert_amount, expert_access, rel_widths = c(1.2, 1))
```

|       The above examples show that, as the regulation-based indicators or composite measures, expert evaluations are not valid proxies for the public funding mechanism. Hence, the question is whether one can identify better proxies that can be used by scholars in comparative research assessing the role of public funding in politics. In fact, there is one measure, labelled as state dependency [@casalbértoa2014]. It reflects the share of state funding in the structure of party income. The higher the share of public funding, the higher the rate of state dependency. Considering that it is a measure derived from real data on public funding, it should be a better measure. While this is true, state dependency might still be a biased measure compared to the actual amount of subsidies per vote. Why so? The main reason is that the the amount of public funding could affect its own share in the structure of party income. The higher the level of state financial support to parties, the lower the incentives to seek private contributions. Moreover, state dependency could still hide considerable variation in public funding per vote despite similar shares in the structure of party income across countries.
|       To test whether this is the case, one needs both measures: subsidies per vote and the state dependency rate. Unfortunately, I do not have data on the state dependency rate for post-communist regimes, given the lack of reliable data on private financing. Therefore, I borrow data from other cross-national studies that provide both indicators [@biezen2017, pp. 87, 90; @casas-zamora2005, p. 49, @nassmacher2009, p. 143]. The relationship between subsidies per vote and state dependency, presented in @fig-dependency, depicts a mixed picture.[^2] Although subsidies per vote and state dependency are positively associated, the strength of the relationship is not sufficient to make them exchangeable.

[^2]:
    | You can replicate @fig-dependency by using this data [{{< fa file-csv >}}](dpf_dependence.csv)`dpf_dependence.csv` along with the R code.

```{r}
#| label: fig-dependency
#| fig-cap: "Relationship between the level of DPF per voter and the share of DPF in the structure of party income"
#| fig-width: 8
#| fig-height: 4
#| 
dpfrv_dependence <- read_csv("dpf_dependence.csv")

### Formula for including regression model on the plot
xy_formula <- y ~ x

### Figure 5 - Left-hand panel: Bivariate relationship between DPFRV vs. state dependence rate, raw data

figure_5_A <- dpfrv_dependence %>%
  ggplot(aes(x = dpf_voter, dpf_percent)) +
  geom_point(aes(shape = author, color = author), size = 2) +
  stat_smooth(method = "lm", color = "grey50", fill = "gray80", formula = xy_formula, se = TRUE) +
  geom_text_repel(aes(label = country_id), family = "serif", size = 3, max.overlaps = 20, segment.color = "grey50", segment.size = 0.1) +
  scale_color_igv(name = "", palette = "default") +
  scale_shape_manual("", values = c(1, 5, 6)) +
  scale_x_continuous(breaks = seq(0,20,3)) +
  stat_regline_equation(
    aes(label =  paste(..eq.label.., ..adj.rr.label.., sep = "~")), label.x = 0.5, label.y = 110, family = "serif", formula = xy_formula
  ) +
  stat_cor(method = "pearson", label.x = 0.5, label.y = 104, family = "serif") +
  custom_theme +
  theme(legend.position = c(0.75, 0.15),
        title = element_text(size = 10),
        legend.key = element_blank(),
        legend.title = element_blank(),
        legend.background = element_blank()) +
  labs(title = "A: Funding per voter vs state dependency rate", x = "Funding per voter", y = "% of subsidies in the structure of party income")

### Figure 5 - Right-hand panel: Bivariate relationship between DPFRV vs. state dependence rate, transformed data

figure_5_B <- dpfrv_dependence %>%
  ggplot(aes(x = dpf_voter, dpf_percent)) +
  geom_point(aes(shape = author, color = author), size = 2) +
  stat_smooth(method = "lm", color = "grey50", fill = "gray80", formula = xy_formula, se = TRUE) +
  geom_text_repel(aes(label = country_id), family = "serif", size = 3, max.overlaps = 20, segment.color = "grey50", segment.size = 0.1) +
  scale_color_igv(name = "", palette = "default") +
  scale_shape_manual("", values = c(1, 5, 6)) +
  scale_x_log10(breaks = c(0,1,2,5,10, 20)) +
  stat_regline_equation(
    aes(label =  paste(..eq.label.., ..adj.rr.label.., sep = "~")), label.x = -1.1, label.y = 95, family = "serif", formula = xy_formula
  ) +
  stat_cor(method = "pearson", label.x = -1.1, label.y = 89, family = "serif") +
  custom_theme +
  theme(legend.position = c(0.75, 0.15),
        title = element_text(size = 10),
        legend.key = element_blank(),
        legend.title = element_blank(),
        legend.background = element_blank()) +
  labs(title = "B: Funding per voter (log10) vs state dependency rate", x = "Funding per voter (log10 transformation)", y = "")

### Combine left-hand and right-hand panels
plot_grid(figure_5_A, figure_5_B)
```

|       As one could notice, expert-based indices are poor substitutes to objective indicators of political financing. While much better, still the degree of state dependence is not sufficiently precise to be used as a substitute to the actual level of public funding. This evidence warns against the blind employment of such measurements to study political finance from a comparative perspective.

::: {#references}
:::
